### 2025-09-18 paper/sec/introduction.tex
Trusted infrastructure.

Making sense of and verifying data-driven claims is hard, even when the evidence base (data sets and related
analyses) is in the public domain. Expert reviewers often struggle to evaluate empirical claims without author
-supplied annotations that provide access to underlying data~\citep{weber20}; paper retractions are frequently
attributable to (non-fraudulent but significant) errors in data management and analysis~\citep{hu25}. The
ability of LLMs to assist with these challenges has received considerable attention recently, for example in
scientific fact-checking~\citep{abu-ahmad25} and figure interpretation~\citep{roberts24}.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} attempts to
address this issue through software infrastructure which transparently connections charts and figures to their
data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.} This enables visual outputs
to provide interactive \emph{provenance queries}, user interactions (e.g. mousing over a visual element) that
allow readers to explore how visual features relate to data. The key value proposition of these approaches is
that the interactions are provided automatically via an underlying programming language (query language or
general-purpose language) which tracks dependencies as data flows through a computation.

@@

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example (top); counterfactual behaviour (bottom)}\label{fig:scigen-example-website}
\end{figure}

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain, because of the difficulty in tracing specific claims back to the data. Peer review of empirical claims
without author-supplied traceability to the underlying data can be challenging~\citep{weber20}; paper
retractions are often attributable to (non-fraudulent but significant) errors in data management and
analysis~\citep{hu25}. The use of LLMs to assist with these challenges has received considerable attention
recently, on problems ranging from fact-checking~\citep{abu-ahmad25} to interpretation of charts and
figures~\citep{roberts24}, LLM interfaces do not support direct interrogation of visual outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} approaches the
transparency problem from this direction, through software infrastructure which transparently connections
charts and figures to their data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.}
This enables visual outputs to provide interactive \emph{provenance queries}, user interactions (e.g. mousing
over a visual element) that allow readers to explore how visual features relate to data. The value proposition
of these approaches is that the interactions are provided automatically via trusted infrastructure, namely an
underlying programming language (say query language or general-purpose language) which tracks dependencies as
data flows through a computation. However, these approaches are limited to outputs computed from data, such as
visualisations.

@@

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example, showing data-driven natural language (top
    vs.~bottom)}\label{fig:scigen-example-website}
\end{figure}
